Impact of grids and dycores in CESM2.2 on the meteorology and climate of the Arctic

### 2021-03-10 arcticgrids.tex
Add files via upload

General Circulation Models (GCMs) are a powerful tool for understanding the meteorology and climate of the poles, which are among the most sensitive regions on Earth to global and environmental change. Despite their importance, numerically, not all GCMs treat the polar regions the same way as they do elsewhere in the domain due to the \textit{pole-problem} \cite{W2007JMSJ}. The wide variety of grids structures and numerical techniques used to represent polar regions in conventional GCMs are evaluated in this study for their impacts on the meteorology and climate of the Arctic.

The pole-problem refers to instability arising from the convergence of meridians into a polar singularity on latitude-longitude grids. In the 1970's this issue was largely defeated through wide-spread adaption of the spectral transform methods in GCMs, which transforms grid point fields into an isotropic representation in wave space. But as computing power has increased, local numerical methods have become more desirable for their ability to run efficiently on massively parallel systems. The pole-problem has thus re-emerged in contemporary climate models using latitude-longitude grids, and some combination of reduced grids and polar filters are necessary to subdue this instability \cite{JW2010LNCSE}.

An alternative approach is to use unstructured grids. Unstructured grids permit quasi-uniform grid spacing globally, thereby eliminating the pole-problem. They allow for more flexible grid-structures than latitude-longitude grids, and can support regional grid refinement that may be used to capture higher resolution in the polar regions. Regional refinement of polar regions may be desirable as latitude-longitude grids, by virtue of the convergence of meridians, have higher horizontal resolution in polar regions compared to quasi-uniform grids containing the same degrees of freedom.

The meteorology and climate of the Arctic is charcterized by a range of processes and scales that are difficult to represent in GCMs \cite{BETAL2001MWR,SG2017MWR,VETAL2018TC}. The Arctic geography consists of a polar ocean surrounded by the northern coasts of the northern hemisphere continents and the large island of Greenland. Greenland is covered in a kilometers thick ice sheet, and receives a substantial portion of its snowfall from synoptic-scale storms traversing up the eastern sea-board of North America that rain out over its steep margins. Intense cold season Polar Lows traverse the Arctic's melange of sea-ice and open ocean, inducing gale-speed winds. Large volumes of fresh water drain into the ocean through river outlets in Canada and Russia, iceberg discharge and melt water runoff emanating from the margins of GrIS. As such, the Arctic serves as a test-bed for comparing the simulated meteorology and climate to different numerical treatments of polar regions.

The atmospheric component of the Community Earth System Model, version 2.2 (CESM; \url{http://www.cesm.ucar.edu/models/cesm2/}), the Community Atmosphere Model, version 6.3 (CAM; \url{https://github.com/ESCOMP/CAM/wiki}), supports a diverse number of atmospheric dynamical cores (hereafter referred to as \textit{dycores}). These range from dycores using latitude-longitude grids, the finite-volume \cite<FV;>{L2004MWR} and eulerian spectral transform \cite<EUL;>{CETAL006JC} models, and dycores built on unstructured grids, the spectral-element \cite<SE;>{LetAl2018JAMES} and finite-volume 3 \cite<FV3;>{PL2007JCP} models. The EUL dycore is the oldest dycore in CAM, and is the least supported of all the dycores. FV3 is the newest dycore in CESM, but it was not fully incorporated into the model at the time this work commenced, and so is omitted from this study. The authors instead focus on the two most well supported dycores in CESM, SE and FV. The goal of this study is to characterize the representation of polar regions using the SE and FV dycores, as these models treat the high-latitudes, e.g., the pole-problem, in very different ways.

@@

>>>>

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the poles, which are among the most sensitive regions on Earth to global and environmental change. Despite their importance, the numerical treatment of polar regions is handled in vastly-different ways due to the \textit{pole-problem} \cite{W2007JMSJ}. The pole-problem refers to instability arising from the convergence of meridians at the polar points on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a). Depending on the numerical design, methods exist to stabilize the pole problem, and latitude-longitude grids may be advantageous for polar processes as structures can be represented with more degrees of freedom than elsewhere in the computational domain. With the recent trend towards globally uniform unstructured grids, any potential benefits of latitude-longitude grids on polar regions may become a relic of the past. In this study, a spectrum of grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 (CESM; \url{http://www.cesm.ucar.edu/models/cesm2/}), from conventional latitude-longitude grids to unstructured grids with globally uniform grid spacing to unstructured grids with regional refinement over the Arctic, are evaluated to understand the impacts of grids and dycores on the simulated characteristics of the Arctic.

In the 1970's the pole problem was largely defeated through wide-spread adoption of efficient spectral transform methods in GCMs. These methods transforms grid point fields into a global, isotropic representation in wave space, where linear operators in the equation set can be solved for exactly. But as computing power has increased, local numerical methods have become more desirable for their ability to run efficiently on massively parallel systems. The pole-problem has thus re-emerged in contemporary climate models using latitude-longitude grids, and some combination of reduced grids and polar filters are necessary to ameliorate this instability \cite{JW2010LNCSE}. Polar filters are akin to a band-aid; they subdue the growth of unstable modes by introducing additional damping to the solution over polar regions. One might expect that this additional damping reduces the effective resolution such that the resolved scales are similar across the entire domain, but this is unlikely to occur in practice.

An alternative approach is unstructured grids. Unstructured grids allow for more flexible grid structures than latitude-longitude grids, permitting quasi-uniform grid spacing globally that eliminates the pole-problem entirely (e.g., Figure~\ref{fig:uni-grids}c). This grid flexibility also allows for variable-resolution or regional grid refinement. Grids can be developed with refinement over polar regions that may in principle make up for any loss in polar resolution in transitioning away from latitude-longitude grids (e.g., Figure~\ref{fig:vr-grids}), although this comes at the cost of a smaller CFL-limiting time-step in the refined region. But unstructured grids scale more efficiently on parallel systems than latitude-longitude grids, resulting in latitude-longitude grids becoming less common, and unstructured grids more common as computing power continued to increase over time.

The meteorology and climate of the Arctic is characterized by a range of processes and scales that are difficult to represent in GCMs \cite{BETAL2001MWR,SG2017MWR,VETAL2018TC}. Synoptic scale storms are well represented at typical GCM resolutions, but mesoscale Polar Lows are not. These mesoscale systems are prevalent during the cold season, and induce gale-force winds that impact fluxes through the underlying sea-ice/ocean interface. The Arctic also contains the Greenland Ice Sheet (hereafter denoted as \textit{GrIS}). While it sits on the largest island in the world (Greenland), GrIS is not well resolved at a typical 100 km grid resolution. GrIS ablation zones exert a primary control on the mass balance of the ice sheet, but are too narrow to resolve at conventional resolutions. The Arctic is therefore well-suited to understand the impact of different grids and dycores on processes that are marginally resolved at conventional GCM resolutions.

The atmospheric component of CESM2.2, the Community Atmosphere Model, version 6.3 (CAM; \url{https://github.com/ESCOMP/CAM/wiki}), supports a diverse number of atmospheric dynamical cores. These range from dycores using latitude-longitude grids, the finite-volume \cite<FV;>{L2004MWR} and eulerian spectral transform \cite<EUL;>{CETAL006JC} models, and dycores built on unstructured grids, the spectral-element \cite<SE;>{LetAl2018JAMES} and finite-volume 3 \cite<FV3;>{PL2007JCP} models. The EUL dycore is the oldest dycore in CAM, and is the least supported of all the dycores. FV3 is the newest dycore in CESM, but it was not fully incorporated into the model at the time this work commenced, and so is omitted from this study. The authors instead focus on the two most well supported dycores in CESM, SE and FV. The goal of this study is to characterize the representation of polar regions using the SE and FV dycores, as these models treat the high-latitudes, e.g., the pole-problem, in very different ways.

@@

\begin{figure}[t]
\begin{center}
         \includegraphics[width=60mm]{figs/temp_grisres.pdf}
\end{center}
\caption{.}
\label{fig:prect}
\end{figure}
